---
title: "analyses - calculations"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(dplyr)
library(tidyr)
library(ggplot2)
library(mstate)
library(relsurv)
library(reshape2)
```

```{r functions}
### msfit.relsurv2 ####
# when doing the analyses for the paper, CBS used an outdated mstate package -> update one line in msfit.relsurv function from the mstate package. (see comment)
# moreover, the default function does not split transitions to a death state into background/excess if there are no events in that specific transition -> some lines are deactivated in this function to make sure that also these transitions can be split. (see comment)
msfit.relsurv2 <- function (msfit.obj, data, split.transitions, ratetable = relsurv::slopop, 
  rmap, time.format = "days", var.pop.haz = c("fixed", "bootstrap", 
    "both"), B = 10, seed = NULL, add.times, substitution = TRUE, 
  link_trans_ind = FALSE) 
{
  if (!requireNamespace("relsurv", quietly = TRUE)) {
    stop("Package \"relsurv\" needed for this function to work. Please install it.", 
      call. = FALSE)
  }
  trans1 <- NULL
  trans2 <- NULL
  b <- NULL
  if (!missing(rmap)) {
    if (substitution) {
      rmap <- substitute(rmap)
    }
  }
  var.pop.haz <- match.arg(var.pop.haz)
  bootstrap_ind <- FALSE
  if (!missing(add.times)) {
    if (max(add.times) > max(msfit.obj[[1]]$time)) {
      warning("Maximum value in add.times is bigger than maximum time in dataset. Do you really want to extrapolate the hazards? \n")
    }
    add.times <- add.times[!(add.times %in% msfit.obj[[1]]$time)]
    msfit.new <- msfit.obj[[1]][0, ]
    for (tr in unique(msfit.obj[[1]]$trans)) {
      msfit_tmp <- subset(msfit.obj[[1]], trans == tr)
      msfit_tmp_2 <- data.frame(time = add.times, Haz = NA, 
        trans = tr)
      msfit_tmp <- rbind(msfit_tmp, msfit_tmp_2)
      msfit_tmp <- msfit_tmp[order(msfit_tmp$time), ]
      msfit_tmp$Haz <- NAfix(msfit_tmp$Haz, 0)
      msfit.new <- rbind(msfit.new, msfit_tmp)
    }
    msfit.obj[[1]] <- msfit.new
    if (length(msfit.obj) == 3) {
      msfit.new <- msfit.obj[[2]][0, ]
      trans.grid <- unique(msfit.obj[[2]][, c("trans1", 
        "trans2")])
      for (tr in 1:nrow(trans.grid)) {
        msfit_tmp <- subset(msfit.obj[[2]], (trans1 == 
          trans.grid[tr, "trans1"] & trans2 == trans.grid[tr, 
          "trans2"]))
        msfit_tmp_2 <- data.frame(time = add.times, 
          varHaz = NA, trans1 = trans.grid[tr, "trans1"], 
          trans2 = trans.grid[tr, "trans2"])
        msfit_tmp <- rbind(msfit_tmp, msfit_tmp_2)
        msfit_tmp <- msfit_tmp[order(msfit_tmp$time), 
          ]
        msfit_tmp$varHaz <- NAfix(msfit_tmp$varHaz, 
          0)
        msfit.new <- rbind(msfit.new, msfit_tmp)
      }
      msfit.obj[[2]] <- msfit.new
    }
  }
  Haz <- msfit.obj[[1]]
  if (length(msfit.obj) == 3) {
    varHaz <- msfit.obj[[2]]
  }
  Year <- 365.241
  Month <- Year/12
  data_orig <- data
  time.format.orig <- time.format
  if (time.format == "days") {
    if (max(msfit.obj[[1]]$time) < 30) 
      warning("Your max time in the data is less than 30 days. If time is not stored in days, please use argument time.format. \n")
  }
  else if (time.format == "years") {
    if (max(msfit.obj[[1]]$time) > 100) 
      warning("Your max time in the data is more than 100 years. If time is not stored in years, please use argument time.format. \n")
    data$Tstart <- data$Tstart * Year
    data$Tstop <- data$Tstop * Year
    data$time <- data$time * Year
    Haz$time <- Haz$time * Year
    if (exists("varHaz")) 
      varHaz$time <- varHaz$time * Year
    time.format <- "days"
  }
  else if (time.format == "months") {
    if (max(msfit.obj[[1]]$time) > 600) 
      warning("Your max time in the data is more than 600 months. If time is not stored in months, please use argument time.format. \n")
    data$Tstart <- data$Tstart * Month
    data$Tstop <- data$Tstop * Month
    data$time <- data$time * Month
    Haz$time <- Haz$time * Month
    if (exists("varHaz")) 
      varHaz$time <- varHaz$time * Month
    time.format <- "days"
  }
  else {
    stop("Argument time.format should take values in c('days', 'years', 'months').")
  }
  Haz_new <- Haz[0, ]
  if (length(msfit.obj) == 3) {
    trans <- msfit.obj[[3]]
  }
  else {
    trans <- msfit.obj[[2]]
  }
  if (missing(split.transitions)) 
    stop("Please define split.transitions.")
  else {
    if (class(split.transitions) == "numeric") {
      if (!all(split.transitions %in% 1:max(trans, na.rm = TRUE))) 
        stop("Invalid transitions used inside argument split.transitions.")
    }
    else stop("Argument split.transitions expects values of class numeric")
    for (i in split.transitions) {
      tmp_state <- which(trans == i, arr.ind = TRUE)[2]
      if (!all(is.na(trans[tmp_state, ]))) 
        stop("You've listed an intermediate transition for which the hazard would have to be split in excess and population hazard. Population mortality tables for intermediate events haven't been implemented in this function. Please include only transitions that go to death states in the split.transitions argument.")
    }
  }
  trans_new <- modify_transMat(trans, split.transitions)
  transitions <- which(!is.na(trans), arr.ind = TRUE)
  link_trans <- list()
  for (i in 1:nrow(transitions)) {
    trans_1 <- trans[transitions[i, 1], transitions[i, 2]]
    if (!(trans_1 %in% split.transitions)) {
      trans_2 <- trans_new[rownames(trans)[transitions[i, 
        1]], colnames(trans)[transitions[i, 2]]]
      link_trans[[trans_1]] <- trans_2
      df_tmp <- msfit.obj[[1]][msfit.obj[[1]]$trans == 
        trans_1, ]
      df_tmp$trans <- trans_2
      Haz_new <- rbind(Haz_new, df_tmp)
    }
    else {
      trans_2 <- c(trans_new[rownames(trans)[transitions[i, 
        1]], paste0(colnames(trans)[transitions[i, 2]], 
        ".p")], trans_new[rownames(trans)[transitions[i, 
        1]], paste0(colnames(trans)[transitions[i, 2]], 
        ".e")])
      link_trans[[trans_1]] <- trans_2
      df_p <- Haz[Haz$trans == trans_1, ]
      df_e <- Haz[Haz$trans == trans_1, ]
      df_subset <- data[(data$from == transitions[i, 1]) & 
        (data$to == transitions[i, 2]), ]
      wh_jump <- which.max(df_e$Haz > 0)
      is_jump <- any(df_e$Haz > 0)
      # these lines below are inactivated so that also death transitions without any events can be split into excess and background
      # if (!is_jump) {
      #   if (!(link_trans_ind == TRUE & substitution == 
      #     FALSE)) {
      #     stop(paste0("There are no events occurring in transition ", 
      #       trans_1, ". Please remove it from the split.transitions argument."))
      #   }
      # }
      if (nrow(df_subset) == 0 | nrow(df_p) == 0) {
        if (nrow(df_p) > 0) {
          df_p$trans <- trans_2[1]
          df_e$trans <- trans_2[2]
        }
      }
      else {
        Hazs <- haz_function(Surv(Tstop, status) ~ 1, 
          data = df_subset, ratetable = ratetable, rmap = rmap, 
          add.times = df_p$time, include.all.times = FALSE)
        wh <- which(Hazs$time %in% (df_p$time))
        wh_l <- c(NA, wh[1:(length(wh) - 1)]) + 1
        wh_l[1] <- 1
        haz.pop <- sapply(1:length(wh), function(x) sum(Hazs$haz.pop[wh_l[x]:wh[x]]))
        haz.excess <- sapply(1:length(wh), function(x) sum(Hazs$haz.excess[wh_l[x]:wh[x]]))
        df_p$trans <- trans_2[1]
        df_p$Haz <- cumsum(haz.pop)
        df_e$trans <- trans_2[2]
        df_e$Haz <- df_e$Haz - df_p$Haz
        old_times <- msfit.obj[[1]]$time[msfit.obj[[1]]$trans == 
          trans_1]
        df_p$time <- old_times
        df_e$time <- old_times
        if (wh_jump < nrow(df_e) & substitution & any(df_e$Haz[(wh_jump + 
          1):nrow(df_e)] < 0)) {
          warning(paste0("The relative survival assumption that the observed hazard can be split in population and excess components might not hold for transition ", 
            trans_1, ". Please check the estimated hazards. Consider removing transition ", 
            trans_1, " from the split.transitions argument. \n"))
        }
      }
      Haz_new <- rbind(Haz_new, df_p, df_e)
    }
  }
  ordering <- order(Haz_new[, "trans"])
  Haz_new <- Haz_new[ordering, , drop = FALSE]
  rownames(Haz_new) <- 1:nrow(Haz_new)
  if (length(msfit.obj) == 3) {
    if (sum(var.pop.haz %in% c("fixed", "bootstrap", "both")) == 
      0) {
      stop("Argument var.pop.haz should take value 'fixed', 'bootstrap' or 'both'.")
    }
    if (var.pop.haz == "fixed" | var.pop.haz == "both") {
      varHaz_new <- varHaz.fixed(varHaz, link_trans, varHaz_original = msfit.obj[[2]])
    }
    if (var.pop.haz == "bootstrap" | var.pop.haz == "both") {
      set.seed(seed)
      all_times <- Haz_new[, c("trans", "time")]
      colnames(all_times)[2] <- "Tstop"
      haz_boot <- msboot.relsurv(theta = msboot.relsurv.boot, 
        data = data_orig, id = colnames(data_orig)[1], 
        B = B, verbose = 0, transmat = trans, all_times = all_times, 
        split.transitions = split.transitions, rmap = rmap, 
        time.format = time.format.orig, boot_orig_msfit = FALSE, 
        ratetable = ratetable, add.times = add.times)
      haz_boot_df <- do.call(rbind.data.frame, haz_boot)
      no_samples <- table(unique(haz_boot_df[, c("trans", 
        "b")])$trans)
      if (any(no_samples < 2) | (!identical(as.integer(names(no_samples)), 
        sort(as.integer(stats::na.omit(as.vector(trans_new))))))) {
        stop("Not enough bootstrap samples so that the variances can be estimated. Please increase value in argument B.")
      }
      haz_boot2 <- aggregate(. ~ time + trans, data = subset(haz_boot_df, 
        select = -b), FUN = stats::var)
      colnames(haz_boot2)[3] <- "varHaz"
      haz_boot3 <- haz_boot2
      haz_boot3$trans1 <- haz_boot3$trans
      colnames(haz_boot3)[2] <- "trans2"
      haz_boot3 <- haz_boot3[, c("time", "varHaz", "trans1", 
        "trans2")]
      if (var.pop.haz == "both") 
        varHaz_new <- rbind(data.frame(varHaz_new, var.pop.haz = "fixed"), 
          data.frame(haz_boot3, var.pop.haz = "bootstrap"))
      else varHaz_new <- haz_boot3
      bootstrap_ind <- TRUE
    }
    ordering <- order(as.vector(as.matrix(varHaz_new[, c("trans1", "trans2")]))) # changed based on version 0.3.3 (CBS currently only has mstate 0.3.2)
    varHaz_new <- varHaz_new[ordering, , drop = FALSE]
    varHaz_new <- stats::na.omit(varHaz_new)
    rownames(varHaz_new) <- 1:nrow(varHaz_new)
  }
  if (length(msfit.obj) == 3) {
    res <- list(Haz = Haz_new, varHaz = varHaz_new, trans = trans_new)
  }
  else {
    res <- list(Haz = Haz_new, trans = trans_new)
  }
  if (link_trans_ind) 
    res$link_trans <- link_trans
  if (bootstrap_ind) 
    res$Haz.boot <- haz_boot
  class(res) <- "msfit"
  return(res)
}


### calculate probtrans ####
# helper functions to combine all necessary steps to calculate hazards and probabilities (data preparation, cox, msfit and probtrans)

# landmark analyses (will be called in func_mstate()):
LM_mstate <- function(msdata, LM, from, min, relsurv = F, split.transitions,
                      time.format, var.pop.haz, ratetable) {
  ms <- cutLMms(msdata, LM = LM) # cut msprep object on landmark time
  store <- vector("list", length(from))
  names(store) <- paste0("from_", from)
  for (i in from) { # repeat analysis from each relevant state
    from_name <- rownames(attr(ms, "trans"))[i]
    message(paste("      from:", i))
    ms2 <- ms %>% # select all persons at risk
      group_by(across(1)) %>%
      filter(any(from == i & Tstart == LM)) %>%
      ungroup() 
    
    # save accessible death states
    death_states <- which(rownames(attr(msdata, "trans")) %in% rownames(attr(msdata, "trans"))[grepl("death", rownames(attr(msdata, "trans")))])
    death_states <- death_states[death_states %in% ms2$to]
    
    ms2 <- ms2 %>%
      filter(trans %in% trans[status == 1 | to %in% death_states]) %>% # keep only transitions with at least 1 event OR if 'to' is in death_states
      mutate(
        Tstart_old = Tstart, Tstop_old = Tstop,
        Tstart = Tstart - LM, Tstop = Tstop - LM # recalculate Tstart and Tstop values
      )
    n <- nrow(unique(ms2[, 1]))
    if (n >= min) { # LM: create new transition matrix excluding unused states/transitions (unless they are accessible death states/transitions)
      tmat_old <- attr(ms, "trans")
      used_states <- ms2 %>%
        filter(status == 1 | to %in% death_states) %>%
        count(from, to, trans) %>%
        pivot_longer(cols = c(from, to)) %>%
        arrange(value) %>%
        pull(value) %>%
        unique()
      
      # create new transition matrix based on which states/transitions are still present
      tmat_new <-
        tra2trans(!(is.na(tmat_old)) &
          tmat_old %in% ms2$trans[ms2$status == 1 | ms2$to %in% death_states])[used_states, used_states]
      rownames(tmat_new) <- rownames(tmat_old)[sort(unique(c(ms2$from, ms2$to)))]
      colnames(tmat_new) <- colnames(tmat_old)[sort(unique(c(ms2$from, ms2$to)))]
      ms2 <- ms2 %>% rename(from_old = from, to_old = to, trans_old = trans)
      ms2 <- ms2 %>% # replace from, to and trans
        left_join(data.frame(trans_old = sort(unique(ms2$trans_old))) %>%
          mutate(trans = 1:n()), by = "trans_old") %>%
        left_join(data.frame(from_old = sort(unique(c(ms2$from_old, ms2$to_old)))) %>%
          mutate(from = 1:n()), by = "from_old") %>%
        left_join(data.frame(to_old = sort(unique(c(ms2$from_old, ms2$to_old)))) %>%
          mutate(to = 1:n()), by = "to_old")
      attr(ms2, "trans") <- tmat_new # replace transition matrix
      from_new <- unique(ms2$from[ms2$from_old == i]) # get new 'from' variable
      
      # calculate cox and msfit objects (currently only non-parametric)
      cox <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data = ms2, method = "breslow")
      fit <- msfit(cox, vartype = "greenwood", trans = tmat_new)
      if (relsurv == T) { # msfit.relsurv2 in case relative survival is needed
        split_transitions <- melt(tmat_new) %>% # get the new split transition numbers
          left_join(melt(tmat_old) %>%
            filter(!is.na(value) & value %in% split.transitions) %>%
            rename(split = value), by = c("Var1" = "from", "Var2" = "to")) %>%
          filter(!is.na(value) & !is.na(split)) %>%
          select(value) %>%
          pull()
        if (length(split_transitions) > 0) {
          mod <- msfit.relsurv2( 
            msfit.obj = fit,
            data = ms2 %>%
              mutate(
                year = year + LM,
                age = age + LM
              ),
            split.transitions = split_transitions,
            ratetable = ratetable,
            time.format = time.format, var.pop.haz = var.pop.haz
          )
          # calculate probabilities
          prob <- probtrans(mod,
            predt = 0, direction = "forward",
            method = "greenwood", covariance = T
          )
        }
        store[[paste0("from_", i)]] <- list(LM, n, mod, prob, i, from_new, from_name)
        names(store[[paste0("from_", i)]]) <-
          c("LM", "n", "mod", "prob", "from_old", "from_new", "from_old_name")
      } else { # calculate probabilities without relative survival
        prob <- probtrans(fit,
          predt = 0, direction = "forward",
          method = "greenwood", covariance = T
        )
        store[[paste0("from_", i)]] <- list(LM, n, fit, prob, i, from_new, from_name)
        names(store[[paste0("from_", i)]]) <-
          c("LM", "n", "msf", "prob", "from_old", "from_new", "from_old_name")
      }
    } else { # in case not enough persons at risk
      store[[paste0("from_", i)]] <- list(LM, n, "from_old_name")
      names(store[[paste0("from_", i)]]) <- c("LM", "n", from_name)
    }
  }
  return(store)
}

# main helper function (it saves an .RDS file with all necessary information (hazards, probabilities, from which states and landmark times the predictions are made, etc.))
func_mstate <- function(msdata, trans = NA,
                        strata = NA, total = T,
                        LM = 0, LM_type = "mstate", # options: both, mstate, LM
                        from = 1, min = 40,
                        survival_type = "observed", # options: both, observed, relsurv
                        ratetable = NA, ratetable_name = NA, split.transitions = NA,
                        time.format = "days", var.pop.haz = "fixed", suffix = "") {
  if (is.null(attr(msdata, "trans"))) {
    if (any(!is.na(trans))) {
      attr(msdata, "trans") <- trans
    } else {
      stop("trans missing in msdata and trans argument")
    }
  }
  if (survival_type %in% c("both", "relsurv") & any(is.na(split.transitions))) {
    stop("relsurv requested but split.transitions missing")
  }
  if (survival_type %in% c("both", "relsurv") & is.na(ratetable_name)) {
    stop("relsurv requested but ratetable_name missing")
  }
  # set data structure for subset analyses
  if (any(!is.na(strata))) {
    df_subset <- msdata %>%
      select(all_of(strata)) %>%
      pivot_longer(everything(), names_to = "strata", values_to = "subset") %>%
      unique()
  } else {
    df_subset <- data.frame(strata = character(), subset = character())
  }
  if (total == T) {
    df_subset <- df_subset %>%
      add_row(strata = "total", subset = "")
  }
  # start calculations
  for (k in 1:nrow(df_subset)) {
    message(paste("calculating for:", df_subset$strata[k], df_subset$subset[k]))
    if (df_subset$strata[k] == "total") {
      msdata2 <- msdata
    } else {
      msdata2 <- msdata[msdata[, df_subset$strata[k]] == df_subset$subset[k], ]
    }
    if (LM_type != "LM" | 0 %in% LM) { # non-landmark analyses (from time = 0 or prediction)
      # calculate transition-specific Cox models
      cox <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans),
        data = msdata2, method = "breslow"
      )
      # calculate cumulative hazards for each transition
      msf <- msfit(cox, vartype = "greenwood", trans = attr(msdata2, "trans"))
      if (survival_type %in% c("both", "relsurv")) { 
        # split death transitions into background/excess using relative survival
        mod <- msfit.relsurv2( 
          msfit.obj = msf, data = msdata2,
          split.transitions = split.transitions,
          ratetable = ratetable,
          time.format = time.format,
          var.pop.haz = var.pop.haz
        )
      }
      # calculate probabilities
      prob <- vector("list", length = ifelse(LM_type == "mstate", length(LM), 1))
      names(prob) <- paste0("time", ifelse(LM_type == "mstate", LM, 0))
      prob_relsurv <- vector("list", length = ifelse(LM_type == "mstate", length(LM), 1))
      names(prob_relsurv) <- paste0("time", ifelse(LM_type == "mstate", LM, 0))
      for (t in LM) {
        if (t == 0 | LM_type != "LM") {
          message(paste("   predt:", t))
          if (survival_type %in% c("both", "observed")) {
            prob[[paste0("time", t)]] <-
              probtrans(msf, predt = t, method = "greenwood", covariance = T)
          }
          if (survival_type %in% c("both", "relsurv")) {
            prob_relsurv[[paste0("time", t)]] <-
              probtrans(mod, predt = t, method = "greenwood", covariance = T)
          }
        }
      }
      # save objects
      if (survival_type %in% c("both", "observed")) {
        saveRDS(
          list(haz = msf, prob = prob),
          paste0(
            "list_mstate", "_",
            gsub(
              "total_", "total",
              gsub(
                " |-", "_",
                gsub(
                  "\\+", "plus",
                  gsub(
                    "<", "below",
                    gsub(
                      ">", "above",
                      paste(
                        df_subset$strata[k],
                        df_subset$subset[k]
                      )
                    )
                  )
                )
              )
            ),
            ifelse(suffix != "", paste0("_", suffix), ""), ".rds"
          )
        )
      }
      if (survival_type %in% c("both", "relsurv")) {
        saveRDS(
          list(haz = mod, prob = prob_relsurv),
          paste0(
            "list_mstate_relsurv_", ratetable_name, "_",
            gsub(
              "total_", "total",
              gsub(
                " |-", "_",
                gsub(
                  "\\+", "plus",
                  gsub(
                    "<", "below",
                    gsub(
                      ">", "above",
                      paste(
                        df_subset$strata[k],
                        df_subset$subset[k]
                      )
                    )
                  )
                )
              )
            ),
            ifelse(suffix != "", paste0("_", suffix), ""), ".rds"
          )
        )
      }
    }
    if (LM_type != "mstate" & any(LM > 0)) { # use LM_mstate for landmark analyses
      LMlist <- vector("list", length = length(LM[LM != 0]))
      names(LMlist) <- paste0("time", LM[LM != 0])
      if (survival_type %in% c("both", "observed")) {
        for (t in LM[LM != 0]) {
          message(paste("   LM observed:", t))
          LMlist[[paste0("time", t)]] <-
            LM_mstate(msdata = msdata2, LM = t, from = from, min = min, relsurv = F)
        }
        saveRDS(
          LMlist,
          paste0(
            "list_mstate_",
            gsub(
              "total_", "total",
              gsub(
                " |-", "_",
                gsub(
                  "\\+", "plus",
                  gsub(
                    "<", "below",
                    gsub(
                      ">", "above",
                      paste(df_subset$strata[k], df_subset$subset[k])
                    )
                  )
                )
              )
            ),
            ifelse(suffix != "", paste0("_", suffix), ""), "_LM.rds"
          )
        )
      }
      if (survival_type %in% c("both", "relsurv")) {
        for (t in LM[LM != 0]) {
          message(paste("   LM relsurv:", t))
          LMlist[[paste0("time", t)]] <-
            LM_mstate(
              msdata = msdata2, LM = t, from = from, min = min, relsurv = T,
              split.transitions = split.transitions, ratetable = ratetable,
              time.format = time.format, var.pop.haz = var.pop.haz
            )
        }
        saveRDS(
          LMlist,
          paste0(
            "list_mstate_relsurv_", ratetable_name, "_",
            gsub(
              "total_", "total",
              gsub(
                " |-", "_",
                gsub(
                  "\\+", "plus",
                  gsub(
                    "<", "below",
                    gsub(
                      ">", "above",
                      paste(
                        df_subset$strata[k],
                        df_subset$subset[k]
                      )
                    )
                  )
                )
              )
            ),
            ifelse(suffix != "", paste0("_", suffix), ""), "_LM.rds"
          )
        )
      }
    }
  }
}
```

```{r data}
data_2020_2021 <- readRDS("data_2020_2021.rds") %>%
  mutate(
    sex = ifelse(GBAGESLACHT == "Mannen", "male", "female"),
    age = age_1Jan20 * 365.241 - 1, # age on 31 December 2019 (time 0)
    age_group = case_when(
      age_1Jan20 < 65 ~ "<65",
      age_1Jan20 < 70 ~ "65-69",
      age_1Jan20 < 80 ~ "70-79",
      age_1Jan20 < 90 ~ "80-89",
      T ~ "90+"
    ),
    sex_age = paste(sex, age_group)
  )

levels_subset <- c(
  "female <65", "female 65-69", "female 70-79", "female 80-89", "female 90+",
  "male <65", "male 65-69", "male 70-79", "male 80-89", "male 90+"
)
```

# total excess mortality
For probability of total excess mortality, use msfit.relsurv2() instead of cmp.rel(), since we have delayed entry for immigrants (not yet implemented in cmp.rel())
- scale is in days
- year is start date (31 December 2019 or 31 May 2020 (so that 1 January and 1 June are time 1))
- sex is sex
- age is age_1Jan*365.241 - 1 (must be in days)
- for 2-year analyses: use same background hazards (already implemented in the ratetables)

```{r total excess mortality - preparation}
# probtrans (needed to translate hazards of msfit.relsurv2 to probabilities) requires at least 2 competing states
# treating censored patients as having an event doesn't matter if it only occurs on the last day of follow-up
tmat_srv <- transMat(
  list(
    c(2, 3),
    c(),
    c()
  ),
  names = c("alive", "death", "censor")
)
tmat_srv

# prepare data in msprep format (needed for mstate)
ms_srv <- data_2020_2021 %>%
  mutate(from = 1, to = 2, trans = 1, time = srv - entry) %>%
  rename(Tstart = entry, Tstop = srv, status = srv_s) %>%
  select(RINPERSOON, from, to, trans, Tstart, Tstop, time, status, sex, age_1Jan20, sex_age)
head(ms_srv)
ms_srv2 <- rbind(
  ms_srv,
  ms_srv %>% mutate(to = 3, trans = 2, status = ifelse(status == 0 & Tstop == 731, 1, 0))
  ) %>%
  arrange(RINPERSOON, trans)
attr(ms_srv2, "trans") <- tmat_srv
class(ms_srv2) <- c("msdata", "data.frame")
events(ms_srv2)
table(data_2020_2021$srv_s) # OK

# get numbers at risk
df_ms_counts_srv <- ms_srv2 %>%
  mutate(
    LM_0 = ifelse(Tstart <= 0 & Tstop > 0, 1, 0),
    LM_152 = ifelse(Tstart <= 152 & Tstop > 152, 1, 0),
    LM_366 = ifelse(Tstart <= 366 & Tstop > 366, 1, 0)
  ) %>%
  pivot_longer(cols = c(starts_with("LM")), names_to = "start", values_to = "present") %>%
  filter(present == 1) %>%
  mutate(start = as.numeric(gsub("LM_", "", start))) %>%
  select(RINPERSOON, from, sex_age, start) %>%
  unique() %>%
  group_by(start, from, sex_age) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(data.frame(state = rownames(tmat_srv), from = 1:length(rownames(tmat_srv))),
    by = "from"
  ) %>%
  mutate(LM = as.Date("2020-01-01") + start)
saveRDS(df_ms_counts_srv, "df_ms_counts_srv.rds")
head(df_ms_counts_srv)
df_ms_counts_srv %>%
  group_by(start, state) %>%
  summarise(sum = sum(n), .groups = "drop") %>%
  filter(sum >= 10) %>%
  pivot_wider(id_cols = state, names_from = start, values_from = sum)
df_ms_counts_srv %>%
  filter(n >= 10) %>%
  pivot_wider(id_cols = c(sex_age, state), names_from = start, values_from = n)
```

## Estimate pandemic-related mortality using default method: HMD (year: 2019, country: NL)
```{r HMD 2019}
# use readRDS("nedpop2.rds"): adjusted ratetable with hazards of 2019 for the years 2020 and 2021
func_mstate(
  msdata = ms_srv2 %>%
    mutate(
      year = as.Date("2019-12-31"),
      age = age_1Jan20 * 365.241 - 1
    ),
  trans = attr(ms_srv2, "trans"),
  strata = "sex_age",
  LM = c(0, 152, 366), LM_type = "LM", survival_type = "both",
  ratetable = readRDS("nedpop2.rds"), ratetable_name = "nedpop2",
  split.transitions = 1, time.format = "days", var.pop.haz = "fixed", suffix = "srv"
)
unique(warnings())
```

## Estimate pandemic-related mortality using mean HMD values for 2015-2019
```{r HMD 2015-2019}
# use readRDS("nedpop_1519.rds"): adjusted ratetable with mean hazards of 2015-2019 for the years 2020 and 2021
func_mstate(
  msdata = ms_srv2 %>%
    mutate(
      year = as.Date("2019-12-31"),
      age = age_1Jan20 * 365.241 - 1
    ),
  trans = attr(ms_srv2, "trans"),
  strata = "sex_age",
  LM = c(0, 152, 366), LM_type = "LM", survival_type = "relsurv",
  ratetable = readRDS("nedpop_1519.rds"), ratetable_name = "nedpop_1519",
  split.transitions = 1, time.format = "days", var.pop.haz = "fixed", suffix = "srv"
)
unique(warnings())
```

# Estimate pandemic-related mortality using ratetable based on microdata: hazard per sex and age
```{r year}
# use readRDS("nedpop_year.rds"): ratetable with calculated background hazards per sex and age
func_mstate(
  msdata = ms_srv2 %>%
    mutate(
      year = as.Date("2019-12-31"),
      age = age_1Jan20 * 365.241 - 1
    ),
  trans = attr(ms_srv2, "trans"),
  strata = "sex_age",
  LM = c(0, 152, 366), LM_type = "LM", survival_type = "relsurv",
  ratetable = readRDS("nedpop_year.rds"), ratetable_name = "nedpop_year",
  split.transitions = 1, time.format = "days", var.pop.haz = "fixed", suffix = "srv"
)
unique(warnings())
``` 

## Estimate pandemic-related mortality using ratetable based on microdata: per sex, age and month
```{r month}
# use readRDS("nedpop_months.rds"): ratetable with calculated background hazards per sex, age and month
func_mstate(
  msdata = ms_srv2 %>%
    mutate(
      year = as.Date("2019-12-31"),
      age = age_1Jan20 * 365.241 - 1
    ),
  trans = attr(ms_srv2, "trans"),
  strata = "sex_age",
  LM = c(0, 152, 366), LM_type = "LM", survival_type = "relsurv",
  ratetable = readRDS("nedpop_months.rds"), ratetable_name = "nedpop_months",
  split.transitions = 1, time.format = "days", var.pop.haz = "fixed", suffix = "srv"
)
unique(warnings())
``` 

# Multi-state model

## build multi-state model
```{r prepare multi-state model}
# how many infection states are needed?
sort(names(df_infect_vacc2_w))
# after vaccination: infect 1-2, rec 1-2
# before any vaccination: infect 1-3, rec 1-3

# build multi-state model structure
tmat3 <- transMat(
  list(
    c(2, 3, 4),   # 1 alive event-free
    c(5, 6),      # 2 vaccination
    c(7, 8, 9),   # 3 infection 1
    c(),          # 4 death event-free
    c(10, 11),    # 5 infection 1 after vaccination
    c(),          # 6 death after vaccination
    c(8, 12, 13), # 7 recovery after infection 1
    c(5, 14),     # 8 vaccination after infection (day of vaccination, possibly before recovery is reached)
    c(),          # 9 death after infection
    c(15, 16),    # 10 recovery after infection 1 after vaccination
    c(),          # 11 death after infection after vaccination
    c(8, 9, 17),  # 12 infection 2
    c(),          # 13 death after recovery after infection
    c(),          # 14 death after vaccination after infection
    c(11, 18),    # 15 infection 2 after vaccination
    c(),          # 16 death after recovery after infection after vaccination
    c(8, 13, 19), # 17 recovery after infection 2
    c(16, 20),    # 18 recovery after infection 2 after vaccination
    c(8, 9, 21),  # 19 infection 3
    c(),          # 20 infection 3 after vaccination (dummy)
    c(8, 13, 22), # 21 recovery after infection 3
    c()           # 22 infection 4 (dummy)
  ), 
  names = c(
    "alive event-free", "vaccination", "infection 1", "death event-free", 
    "infection 1 after vacc", "death after vaccination", "recovery after infection 1", 
    "vaccination after infection", "death after infection", 
    "recovery after infection 1 after vacc", 
    "death after infection after vacc", "infection 2", 
    "death after recovery after infection", "death after vaccination after infection ", 
    "infection 2 after vacc", "death after recovery after infection after vacc", 
    "recovery after infection 2", "recovery after infection 2 after vacc", 
    "infection 3", "infection 3 after vacc (dummy)", 
    "recovery after infection 3", "infection 4 (dummy)" 
  )
)
tmat3
tmat3 %>%
  as_tibble(rownames = "from") %>%
  pivot_longer(-from, names_to = "to", values_to = "trans") %>%
  filter(!is.na(trans)) %>%
  print(n = Inf) # OK

data_2020_2021 %>%
  select(RINPERSOON, entry, vacc, contains("vacc_infect"), contains("vacc_rec")) %>%
  select(-ends_with("_s")) %>%
  pivot_longer(cols = -c(RINPERSOON, entry)) %>%
  filter(!is.na(value) & value <= entry) %>%
  group_by(name) %>%
  summarise(n = n())

# get starting states:
table(data_2020_2021$start_infect_s, useNA = "ifany")
data_2020_2021 <- data_2020_2021 %>%
  mutate(
    start3_s = ifelse(!is.na(start_infect_s),
      ifelse(grepl("after", start_infect_s) |
        (is.na(date_vaccination) | (!is.na(date_vaccination) & date_vaccination > start)),
      start_infect_s, "vaccination after infection"
      ),
      ifelse(!is.na(date_vaccination) & date_vaccination <= start,
        "vaccination", "alive event-free"
      )
    ),
    start3_s_n = case_when(
      start3_s == "alive event-free" ~ 1,
      start3_s == "vaccination" ~ 2,
      start3_s == "before_vacc_infect_1" ~ 3,
      start3_s == "before_vacc_rec_1" ~ 7,
      start3_s == "vaccination after infection" ~ 8
    )
  )
table(data_2020_2021$start3_s, useNA = "ifany")

data_2020_2021 %>%
  summarise(n(),
    no_vacc = length(which(is.na(date_vaccination))),
    min_int = min(date_vaccination - start, na.rm = T),
    max_int = max(date_vaccination - start, na.rm = T)
  )
# some never had vaccination, the rest had vaccination after start
data_2020_2021 %>%
  filter(start3_s %in% c("alive event-free", "vaccination", "vaccination after infection")) %>%
  group_by(start3_s) %>%
  summarise(n(),
    no_vacc = length(which(is.na(date_vaccination))),
    min_vacc_int = min(date_vaccination - start, na.rm = T),
    max_vacc_int = max(date_vaccination - start, na.rm = T),
    no_infect = length(which(is.na(infect_1))),
    min_infect_int = min(infect_1 - start, na.rm = T),
    max_infect_int = max(infect_1 - start, na.rm = T)
  ) %>%
  print(width = Inf)

data_2020_2021 <- data_2020_2021 %>%
  mutate(
    across(
      c((contains("before_vacc") | contains("after_vacc")) & ends_with("_s")),
      ~ ifelse(is.na(.x), 0, .x)
    ),
    across(
      c((contains("before_vacc") | contains("after_vacc")) & ends_with("_int")),
      ~ ifelse(is.na(.x), srv, .x)
    )
  )

df_infect_vacc %>%
  filter(time == date_vaccination) %>%
  select(state, timing) %>%
  unique()
# none

# create msprep object: too large dataset for msprep -> do in chunks
data_2020_2021$chunk <- rep(seq_len(ceiling(nrow(data_2020_2021) / 99997)),
  each = 99997, length.out = nrow(data_2020_2021)
)
table(data_2020_2021$chunk) # 39 chunks

for (i in 1:max(data_2020_2021$chunk)) {
  saveRDS(
    msprep(
      data = data_2020_2021 %>% filter(data_2020_2021$chunk == i) %>%
        # dummies with full path through model
        add_row(
          RINPERSOON = "dummy1", entry_s = 1, entry = 0,
          before_vacc_infect_1_s = 1, before_vacc_infect_1_int = 1,
          before_vacc_rec_1_s = 1, before_vacc_rec_1_int = 2,
          before_vacc_infect_2_s = 1, before_vacc_infect_2_int = 3,
          before_vacc_rec_2_s = 1, before_vacc_rec_2_int = 4,
          before_vacc_infect_3_s = 1, before_vacc_infect_3_int = 5,
          before_vacc_rec_3_s = 1, before_vacc_rec_3_int = 6,
          vacc_s = "vaccination", vacc = 8,
          after_vacc_infect_1_s = 1, after_vacc_infect_1_int = 9,
          after_vacc_rec_1_s = 1, after_vacc_rec_1_int = 10,
          after_vacc_infect_2_s = 1, after_vacc_infect_2_int = 11,
          after_vacc_rec_2_s = 1, after_vacc_rec_2_int = 12,
          srv_s = 1, srv = 13
        ) %>%
        add_row(
          RINPERSOON = "dummy2", entry_s = 1, entry = 0,
          before_vacc_infect_1_s = 1, before_vacc_infect_1_int = 1,
          before_vacc_rec_1_s = 1, before_vacc_rec_1_int = 2,
          before_vacc_infect_2_s = 1, before_vacc_infect_2_int = 3,
          before_vacc_rec_2_s = 1, before_vacc_rec_2_int = 4,
          before_vacc_infect_3_s = 1, before_vacc_infect_3_int = 5,
          before_vacc_rec_3_s = 1, before_vacc_rec_3_int = 6,
          vacc_s = "vaccination", vacc = 8,
          after_vacc_infect_1_s = 1, after_vacc_infect_1_int = 9,
          after_vacc_rec_1_s = 1, after_vacc_rec_1_int = 10,
          after_vacc_infect_2_s = 1, after_vacc_infect_2_int = 11,
          after_vacc_rec_2_s = 1, after_vacc_rec_2_int = 12,
          srv_s = 1, srv = 13
        ) %>%
        mutate(
          vacc_s = ifelse(vacc_s == "vaccination", 1, 0),
          after_vacc_infect_3_s = 0, before_vacc_infect_4_s = 0
        ),
      trans = tmat3, id = "RINPERSOON",
      time = c(
        "entry", "vacc",
        "before_vacc_infect_1_int", "srv",
        "after_vacc_infect_1_int", "srv",
        "before_vacc_rec_1_int", "vacc",
        "srv", "after_vacc_rec_1_int",
        "srv", "before_vacc_infect_2_int",
        "srv", "srv", "after_vacc_infect_2_int",
        "srv", "before_vacc_rec_2_int",
        "after_vacc_rec_2_int", "before_vacc_infect_3_int",
        "srv", "before_vacc_rec_3_int", "srv"
      ),
      status = c(
        "entry_s", "vacc_s",
        "before_vacc_infect_1_s", "srv_s",
        "after_vacc_infect_1_s", "srv_s",
        "before_vacc_rec_1_s", "vacc_s",
        "srv_s", "after_vacc_rec_1_s",
        "srv_s", "before_vacc_infect_2_s",
        "srv_s", "srv_s", "after_vacc_infect_2_s",
        "srv_s", "before_vacc_rec_2_s",
        "after_vacc_rec_2_s", "before_vacc_infect_3_s",
        "after_vacc_infect_3_s", "before_vacc_rec_3_s", "before_vacc_infect_4_s"
      ),
      start = list(
        state = c(data_2020_2021$start3_s_n[data_2020_2021$chunk == i], 1, 1),
        time = c(data_2020_2021$entry[data_2020_2021$chunk == i], 0, 0)
      )
    ),
    paste0("ms3.", i, ".rds")
  )
}
# combine the chunks to get the full msprep object
ms3 <- bind_rows(lapply(list.files(pattern = "ms3.*.rds")[which(!grepl("June", list.files(pattern = "ms3.*.rds")))][1:39], readRDS)) %>%
  filter(!grepl("dummy", RINPERSOON))

head(ms3)
events(ms3)
saveRDS(ms3, "ms3.rds")

# combine recurrent states and transitions:
tmat3_s <- transMat(
  list(
    c(2, 3, 4),  # 1 alive event-free
    c(5, 6),     # 2 vaccination
    c(7, 8, 9),  # 3 infection
    c(),         # 4 death event-free
    c(10, 11),   # 5 infection after vaccination
    c(),         # 6 death after vaccination
    c(3, 8, 12), # 7 recovery after infection
    c(5, 13),    # 8 vaccination after infection (day of vaccination, possibly before recovery is reached) 
    c(),         # 9 death after infection
    c(5, 14),    # 10 recovery after infection after vaccination
    c(),         # 11 death after infection after vaccination
    c(),         # 12 death after recovery after infection
    c(),         # 13 death after vaccination after infection
    c()          # 14 death after recovery after infection after vaccination
  ), 
  names = c( # use final names:
    "alive event-free", "vaccination", "recent infection", "death event-free", 
    "recent infection after vaccination", "death after vaccination", "non-recent infection", 
    "vaccination after infection", "death after recent infection", 
    "non-recent infection after vaccination", 
    "death after recent infection after vaccination", "death after non-recent infection",
    "death after vaccination after infection",  
    "death after non-recent infection after vaccination"
  )
)

tmat3_s
tmat3_s %>%
  as_tibble(rownames = "from") %>%
  pivot_longer(-from, names_to = "to", values_to = "trans") %>%
  filter(!is.na(trans)) %>%
  print(n = Inf) # OK

# renumber from, to and trans
ms3_s <- ms3 %>%
  mutate(
    from = case_when(
      from %in% c(1:11) ~ from,
      from %in% c(12, 19, 22) ~ 3,
      from %in% c(13, 14) ~ from - 1,
      from %in% c(15, 20) ~ 5,
      from == 16 ~ 14,
      from %in% c(17, 21) ~ 7,
      from == 18 ~ 10
    ),
    to = case_when(
      to %in% c(1:11) ~ to,
      to %in% c(12, 19, 22) ~ 3,
      to %in% c(13, 14) ~ to - 1,
      to %in% c(15, 20) ~ 5,
      to == 16 ~ 14,
      to %in% c(17, 21) ~ 7,
      to == 18 ~ 10
    ),
    trans = case_when(
      trans %in% c(1:10, 13:17) ~ trans,
      trans %in% c(11, 23, 31) ~ 12,
      trans %in% c(12, 25, 33) ~ 11,
      trans %in% c(18, 28) ~ 7,
      trans %in% c(19, 29) ~ 8,
      trans %in% c(20, 30) ~ 6,
      trans == 21 ~ 10,
      trans == 22 ~ 9,
      trans %in% c(24, 32) ~ 13,
      trans == 26 ~ 17,
      trans == 27 ~ 16
    )
  )
attr(ms3_s, "trans") <- tmat3_s

# check from:
cbind(
  ms3 %>% select(from) %>%
    left_join(data.frame(from = c(1:length(rownames(tmat3))), name = rownames(tmat3))),
  ms3_s %>% select(from) %>%
    left_join(data.frame(from = c(1:length(rownames(tmat3_s))), name = rownames(tmat3_s))) %>%
    rename(from_s = from, name_s = name)
) %>%
  unique() %>%
  arrange(name_s, name)
# check to:
cbind(
  ms3 %>% select(to) %>%
    left_join(data.frame(to = c(1:length(rownames(tmat3))), name = rownames(tmat3))),
  ms3_s %>% select(to) %>%
    left_join(data.frame(to = c(1:length(rownames(tmat3_s))), name = rownames(tmat3_s))) %>%
    rename(to_s = to, name_s = name)
) %>%
  unique() %>%
  arrange(name_s, name) 
# check trans:
cbind(
  ms3 %>% select(trans) %>% left_join(tmat3 %>% as_tibble(rownames = "from") %>%
    pivot_longer(-from, names_to = "to", values_to = "trans") %>%
    filter(!is.na(trans)) %>% mutate(name = paste(from, "->", to)) %>%
    select(trans, name)),
  ms3_s %>% select(trans) %>% left_join(tmat3_s %>% as_tibble(rownames = "from") %>%
    pivot_longer(-from, names_to = "to", values_to = "trans") %>%
    filter(!is.na(trans)) %>% mutate(name = paste(from, "->", to)) %>%
    select(trans, name)) %>%
    rename(trans_s = trans, name_s = name)
) %>%
  unique() %>%
  arrange(name_s, name) 

events(ms3_s)

# select period starting from 31 May 2020 (before that there was no reliable testing)
as.numeric(as.Date("2020-05-31") - as.Date("2019-12-31")) # 152 days
ms3_s_June <- ms3_s %>%
  mutate(
    Tstart = Tstart - 152, Tstart = ifelse(Tstart < 0, 0, Tstart),
    Tstop = Tstop - 152
  ) %>%
  mutate(time = Tstop - Tstart) %>%
  filter(Tstop > 0)
nrow(ms3_s_June)

ms3_s_June %>%
  filter(RINPERSOON %in% RINPERSOON[Tstart == max(Tstop)][1])
# since censoring occurs on 31 December 2021, persons can have rows with time 0 (if they also had an infection or a vaccination on 31 December 2021)
ms3_s_June %>%
  filter(time == 0) %>%
  select(from:status) %>%
  unique()
# doesn't occur anywhere else; remove these rows
ms3_s_June <- ms3_s_June %>% filter(time != 0)
nrow(ms3_s_June)

attr(ms3_s_June, "trans") <- tmat3_s
class(ms3_s_June) <- c("msdata", "data.frame")

events(ms3_s_June) # no empty rows -> OK

# add subset variable
if (!"sex" %in% names(ms3_s_June)) {
  ms3_s_June <- ms3_s_June %>% left_join(
    data_2020_2021 %>%
      select(RINPERSOON, sex, age_1Jan20, sex_age),
    by = "RINPERSOON"
  )
}
saveRDS(ms3_s_June, "ms3_s_June.rds")

# order of states
state_levels_ms <- c(
  "death after non-recent infection after vaccination",
  "non-recent infection after vaccination", "death after vaccination after infection",
  "death after non-recent infection", "death after recent infection after vaccination",
  "recent infection after vaccination", "vaccination after infection",
  "non-recent infection", "death after vaccination", "vaccination", "recent infection",
  "death after recent infection", "death event-free", "alive event-free"
)

state_levels_ms_relsurv <- c(
  "death after non-recent infection after vaccination.p",
  "death after non-recent infection after vaccination.e",
  "non-recent infection after vaccination", "death after vaccination after infection.p",
  "death after vaccination after infection.e", "death after non-recent infection.p",
  "death after non-recent infection.e", "death after recent infection after vaccination.p",
  "death after recent infection after vaccination.e", "recent infection after vaccination",
  "vaccination after infection", "non-recent infection", "death after vaccination.p",
  "death after vaccination.e", "vaccination", "recent infection",
  "death after recent infection.p", "death after recent infection.e", "death event-free.p",
  "death event-free.e", "alive event-free"
)

# numbers at risk (present in state)
df_ms_counts <- ms3_s_June %>%
  mutate(
    LM_0 = ifelse(Tstart <= 0 & Tstop > 0, 1, 0),
    LM_214 = ifelse(Tstart <= 214 & Tstop > 214, 1, 0),
    LM_365 = ifelse(Tstart <= 365 & Tstop > 365, 1, 0),
    LM_457 = ifelse(Tstart <= 457 & Tstop > 457, 1, 0),
    LM_548 = ifelse(Tstart <= 548 & Tstop > 548, 1, 0)
  ) %>%
  pivot_longer(cols = c(starts_with("LM")), names_to = "start", values_to = "present") %>%
  filter(present == 1) %>%
  mutate(start = as.numeric(gsub("LM_", "", start))) %>%
  select(RINPERSOON, from, sex_age, start) %>%
  unique() %>%
  group_by(start, from, sex_age) %>%
  summarise(n = n(), .groups = "drop") %>%
  left_join(data.frame(state = rownames(tmat3_s), from = 1:length(rownames(tmat3_s))),
    by = "from"
  ) %>%
  mutate(LM = as.Date("2020-06-01") + start)
saveRDS(df_ms_counts, "df_ms_counts.rds")

head(df_ms_counts)
df_ms_counts %>%
  group_by(start, state) %>%
  summarise(sum = sum(n), .groups = "drop") %>%
  filter(sum >= 10) %>%
  pivot_wider(id_cols = state, names_from = start, values_from = sum)
df_ms_counts %>%
  filter(n >= 10) %>%
  pivot_wider(id_cols = c(sex_age, state), names_from = start, values_from = n)
df_ms_counts %>%
  filter(n >= 10) %>%
  mutate(
    sex_age = factor(gsub("male", "Men", gsub("female", "Women", sex_age)),
      levels = rev(c(
        "Women <65", "Men <65",
        "Women 65-69", "Men 65-69",
        "Women 70-79", "Men 70-79",
        "Women 80-89", "Men 80-89",
        "Women 90+", "Men 90+"
      ))
    ),
    date = start + as.Date("2020-06-01"),
    state = factor(state, levels = c(
      "alive event-free", "vaccination",
      "recent infection after vaccination",
      "recent infection", "vaccination after infection",
      "non-recent infection after vaccination",
      "non-recent infection"
    ))
  ) %>%
  ggplot(aes(x = as.factor(date), y = n, fill = sex_age)) +
  geom_col(color = "black") +
  facet_wrap(~state, scales = "free_y") +
  theme_bw() +
  theme(
    axis.text = element_text(color = "black"),
    legend.title = element_blank(), axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90), panel.grid = element_blank()
  ) +
  scale_x_discrete() +
  scale_fill_viridis_d(option = "H", begin = 0.1, direction = -1) +
  ylab("number of unique persons present in state on landmark time\n(group sizes <10 not shown)")

df_ms_counts %>%
  filter(n >= 10) %>%
  group_by(start, from) %>%
  mutate(total = sum(n), prop = n / total) %>%
  mutate(
    sex_age = factor(gsub("male", "Men", gsub("female", "Women", sex_age)),
      levels = rev(c(
        "Women <65", "Men <65",
        "Women 65-69", "Men 65-69",
        "Women 70-79", "Men 70-79",
        "Women 80-89", "Men 80-89",
        "Women 90+", "Men 90+"
      ))
    ),
    date = start + as.Date("2020-06-01"),
    state = factor(state, levels = c(
      "alive event-free", "vaccination",
      "recent infection after vaccination",
      "recent infection", "vaccination after infection",
      "non-recent infection after vaccination",
      "non-recent infection"
    ))
  ) %>%
  ggplot(aes(x = as.factor(date), y = prop, fill = sex_age)) +
  geom_col(color = "black") +
  facet_wrap(~state) +
  theme_bw() +
  theme(
    axis.text = element_text(color = "black"),
    legend.title = element_blank(), axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90), panel.grid = element_blank()
  ) +
  scale_x_discrete() +
  scale_fill_viridis_d(option = "H", begin = 0.1, direction = -1) +
  ylab("distribution of sex/age categories per state and landmark time\n(group sizes <10 not shown")

df_ms_counts %>%
  filter(n >= 10) %>%
  group_by(start, sex_age) %>%
  mutate(total = sum(n), prop = n / total) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(start, sex_age), names_from = from, values_from = prop) %>%
  print(n = Inf) %>%
  filter(if_any(c(`1`, `3`, `7`, `2`, `5`, `8`, `10`), ~ !is.na(.x) & .x > 0.90))

df_ms_counts %>%
  filter(n >= 10) %>%
  group_by(start, sex_age) %>%
  mutate(total = sum(n), prop = n / total) %>%
  mutate(
    sex_age = factor(gsub("male", "Men", gsub("female", "Women", sex_age)),
      levels = rev(c(
        "Women <65", "Men <65",
        "Women 65-69", "Men 65-69",
        "Women 70-79", "Men 70-79",
        "Women 80-89", "Men 80-89",
        "Women 90+", "Men 90+"
      ))
    ),
    date = start + as.Date("2020-06-01"),
    state = factor(state, levels = c(
      "alive event-free", "vaccination",
      "recent infection after vaccination",
      "recent infection", "vaccination after infection",
      "non-recent infection after vaccination",
      "non-recent infection"
    ))
  ) %>%
  ggplot(aes(x = as.factor(date), y = prop, fill = state)) +
  geom_col(color = "black") +
  facet_wrap(~sex_age) +
  theme_bw() +
  theme(
    axis.text = element_text(color = "black"),
    legend.title = element_blank(), axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 90), panel.grid = element_blank()
  ) +
  scale_x_discrete() +
  scale_fill_viridis_d(option = "H", begin = 0.1, direction = -1) +
  ylab("distribution of sex/age categories per state and landmark time\n(group sizes <10 not shown")

# get at risk numbers for each timepoint
df_ms_atrisk <- ms3_s_June %>%
  select(RINPERSOON, from, Tstart, Tstop, sex_age) %>%
  unique()
df_ms_atrisk <- df_ms_atrisk %>% count(from, Tstart, Tstop, sex_age)
df_ms_atrisk <- df_ms_atrisk %>% mutate(Tstop2 = Tstop - 1)
df_ms_atrisk <- df_ms_atrisk %>%
  rowwise() %>%
  do(data.frame(
    from = .$from, sex_age = .$sex_age, n = .$n,
    time = seq(.$Tstart, .$Tstop2, by = 1)
  ))
head(df_ms_atrisk)
df_ms_atrisk <- df_ms_atrisk %>%
  group_by(from, sex_age, time) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  left_join(
    data.frame(
      from = 1:length(tmat3_s),
      from_name = factor(rownames(tmat3_s), levels = c(rownames(tmat3_s)))
    ),
    by = "from"
  )
```

```{r diagram}
# box diagram showing the structure of the multi-state model
  # states (boxes)
df_states <- data.frame(state = rownames(tmat3_s)) %>%
  mutate(
    xmin = case_when(
      state %in% c("death event-free", "alive event-free") ~ 1,
      state %in% c(
        "death after vaccination", "vaccination", "recent infection",
        "death after recent infection"
      ) ~ 25,
      state %in% c(
        "death after recent infection after vaccination",
        "recent infection after vaccination", "vaccination after infection",
        "non-recent infection"
      ) ~ 50,
      T ~ 75
    ),
    xmax = xmin + 20,
    xmid = xmin + 10,
    ymax = case_when(
      state %in% c(
        "death event-free", "death after vaccination",
        "death after recent infection after vaccination",
        "death after non-recent infection after vaccination"
      ) ~ 100,
      state %in% c(
        "alive event-free", "vaccination", "recent infection after vaccination",
        "non-recent infection after vaccination"
      ) ~ 80,
      state %in% c(
        "recent infection", "vaccination after infection",
        "death after vaccination after infection"
      ) ~ 60,
      T ~ 40
    ),
    ymin = ymax - 10,
    ymid = ymax - 5,
    color = c(
      "#00B0F0", "#00B0F0", "#00B0F0", "#CC0000", "#00B0F0", "#CC0000", "#00B0F0",
      "#00B0F0", "#CC0000", "#00B0F0", "#CC0000", "#CC0000", "#CC0000", "#CC0000"
    )
  ) %>%
  rowwise() %>%
  mutate(fill = colorRampPalette(c(color, "white"))(4)[3]) %>%
  ungroup()
df_states

  # transitions (arrows)
df_trans <- melt(tmat3_s) %>%
  filter(!is.na(value)) %>%
  select(-value) %>%
  left_join(df_states %>% rename(from = state), by = "from") %>%
  left_join(df_states %>% rename(to = state), by = "to") %>%
  mutate(
    x = case_when(
      xmid.x == xmid.y ~ xmid.x,
      xmid.x < xmid.y ~ xmax.x,
      T ~ xmin.x
    ),
    xend = case_when(
      xmid.x == xmid.y ~ xmid.x,
      xmid.x < xmid.y ~ xmin.y,
      T ~ xmax.y
    ),
    y = case_when(
      xmid.x == xmid.y & ymid.x < ymid.y ~ ymax.x,
      xmid.x == xmid.y ~ ymin.x,
      T ~ ymid.x
    ),
    yend = case_when(
      xmid.x == xmid.y & ymid.x < ymid.y ~ ymin.y,
      xmid.x == xmid.y ~ ymax.y,
      T ~ ymid.y
    )
  )
df_trans

# plot diagram
ggplot() +
  geom_segment(
    data = df_trans, aes(x = x, xend = xend, y = y, yend = yend),
    arrow = arrow(length = unit(0.03, "npc"), type = "closed", angle = 20)
  ) +
  geom_rect(
    data = df_states, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "white",
    color = df_states$color, linewidth = 1.5
  ) +
  geom_text(
    data = df_states, aes(x = xmid, y = ymid, label = stringr::str_wrap(state, 20)),
    hjust = 0.5, vjust = 0.5, size = 4
  ) +
  theme_void() +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01))

# add numbers to state names
ggplot() +
  geom_segment(
    data = df_trans, aes(x = x, xend = xend, y = y, yend = yend),
    arrow = arrow(length = unit(0.03, "npc"), type = "closed", angle = 20)
  ) +
  geom_rect(
    data = df_states, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "white",
    color = df_states$color, linewidth = 1.5
  ) +
  geom_text(
    data = df_states %>%
      mutate(state = paste0(c(1, 4, 5, 2, 8, 3, 10, 9, 6, 12, 7, 14, 13, 11), ". ", df_states$state)),
    aes(x = xmid, y = ymid, label = stringr::str_wrap(state, 20)),
    hjust = 0.5, vjust = 0.5, size = 4
  ) +
  theme_void() +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01))
```

## without relsurv and with relsurv based on HMD 2015-2019
```{r mstate without relsurv and HMD 2015-2019}
# calculate hazards and probabilities
func_mstate(
  msdata = ms3_s_June %>%
    mutate(
      year = as.Date("2020-05-31"),
      age = age_1Jan20 * 365.241 - 1 + 152
    ),
  trans = attr(ms3_s_June, "trans"),
  strata = "sex_age",
  LM = c(0, 366 - 152, 517 - 152, 609 - 152, 700 - 152), LM_type = "both",
  from = c(1, 2, 3, 5, 7, 8, 10), survival_type = "both",
  ratetable = readRDS("nedpop_1519.rds"), ratetable_name = "nedpop_1519",
  split.transitions = c(3, 5, 8, 10, 13, 15, 17),
  time.format = "days",
  var.pop.haz = "fixed"
)
unique(warnings())
```

## with relsurv based on HMD 2019
```{r mstate with relsurv based on 2019}
# calculate hazards and probabilities
func_mstate(
  msdata = ms3_s_June %>%
    mutate(
      year = as.Date("2020-05-31"),
      age = age_1Jan20 * 365.241 - 1 + 152
    ),
  trans = attr(ms3_s_June, "trans"),
  strata = "sex_age",
  LM = c(0, 366 - 152, 517 - 152, 609 - 152, 700 - 152), LM_type = "both",
  from = c(1, 2, 3, 5, 7, 8, 10), survival_type = "relsurv",
  ratetable = readRDS("nedpop2.rds"), ratetable_name = "nedpop2",
  split.transitions = c(3, 5, 8, 10, 13, 15, 17),
  time.format = "days",
  var.pop.haz = "fixed"
)
unique(warnings())
```

## with relsurv based on microdata: hazards per sex and age
```{r mstate year}
# calculate hazards and probabilities
func_mstate(
  msdata = ms3_s_June %>%
    mutate(
      year = as.Date("2020-05-31"),
      age = age_1Jan20 * 365.241 - 1 + 152
    ),
  trans = attr(ms3_s_June, "trans"),
  strata = "sex_age",
  LM = c(0, 366 - 152, 517 - 152, 609 - 152, 700 - 152), LM_type = "both",
  from = c(1, 2, 3, 5, 7, 8, 10), survival_type = "relsurv",
  ratetable = readRDS("nedpop_year.rds"), ratetable_name = "nedpop_year",
  split.transitions = c(3, 5, 8, 10, 13, 15, 17),
  time.format = "days",
  var.pop.haz = "fixed"
)
unique(warnings())
```

## with relsurv based on microdata: per sex, age and month
```{r mstate month}
# calculate hazards and probabilities
func_mstate(
  msdata = ms3_s_June %>%
    mutate(
      year = as.Date("2020-05-31"),
      age = age_1Jan20 * 365.241 - 1 + 152
    ),
  trans = attr(ms3_s_June, "trans"),
  strata = "sex_age",
  LM = c(0, 366 - 152, 517 - 152, 609 - 152, 700 - 152), LM_type = "both",
  from = c(1, 2, 3, 5, 7, 8, 10), survival_type = "relsurv",
  ratetable = readRDS("nedpop_months.rds"), ratetable_name = "nedpop_months",
  split.transitions = c(3, 5, 8, 10, 13, 15, 17),
  time.format = "days",
  var.pop.haz = "fixed"
)
unique(warnings())
```

```{r check excess mortality after vaccination}
# vaccination campaign: first frail people, then based on age -> to correctly analyse this, we need to use the same age groups as used in the vaccination campaign:
ms3_s_June2 <- ms3_s_June %>%
  left_join(data_2020_2021 %>% select(RINPERSOON, BIRTHDAY) %>%
    mutate(
      birthyear = as.numeric(substr(BIRTHDAY, 1, 4)),
      birthcohort = case_when(
        birthyear < 1931 ~ "<1931",
        birthyear < 1936 ~ "1931-1935",
        birthyear < 1941 ~ "1936-1940",
        birthyear == 1956 ~ "1956",
        birthyear < 1947 ~ "1941-1946",
        birthyear < 1952 ~ "1947-1951",
        birthyear == 1952 ~ "1952",
        birthyear < 1956 ~ "1953-1955"
      ),
      birthcohort = factor(birthcohort,
        levels = c(
          "<1931", "1931-1935", "1936-1940", "1941-1946",
          "1947-1951", "1952", "1953-1955", "1956"
        )
      )
    ))
ms3_s_June2 %>% select(RINPERSOON, birthcohort, sex_age) %>% unique() %>% 
  count(birthcohort, sex_age)
df_ms_counts_birthcohort <- ms3_s_June2 %>% 
  select(RINPERSOON, birthcohort) %>% unique() %>% 
  count(birthcohort) %>% mutate(n = ifelse(n >= 10, n, NA))
saveRDS(df_ms_counts_birthcohort, "df_ms_counts_birthcohort.rds")

# calculate hazards and probabilities
func_mstate(
  msdata = ms3_s_June2 %>%
    mutate(
      year = as.Date("2020-05-31"),
      age = age_1Jan20 * 365.241 - 1 + 152
    ),
  trans = attr(ms3_s_June2, "trans"),
  strata = "birthcohort", total = F,
  survival_type = "relsurv",
  ratetable = readRDS("nedpop_months.rds"), ratetable_name = "nedpop_months",
  split.transitions = c(3, 5, 8, 10, 13, 15, 17),
  time.format = "days",
  var.pop.haz = "fixed"
)
unique(warnings())
```

# Cumulative incidences
```{r cuminc}
#### infection ####
# with death as competing event
fit_inf_deathComp <- Cuminc("infectJun", "infectJun_s",
  data_2020_2021 %>%
    filter(start <= as.Date("2020-01-01") & infectJun > 152) %>%
    mutate(infectJun = infectJun - 152) %>%
    as.data.frame(),
  failcodes = c("infection", "death")
)
saveRDS(fit_inf_deathComp, "fit_inf_deathComp.rds")

# subset analyses
fit_inf_deathComp_subsets <- Cuminc("infectJun", "infectJun_s",
  data_2020_2021 %>%
    filter(start <= as.Date("2020-01-01") & infectJun > 152) %>%
    mutate(infectJun = infectJun - 152) %>%
    as.data.frame(),
  failcodes = c("infection", "death"),
  group = "sex_age"
)
saveRDS(fit_inf_deathComp_subsets, "fit_inf_deathComp_subsets.rds")

#### vaccination #####
# with death as competing event
fit_vacc_deathComp <- Cuminc("vacc", "vacc_s",
  data_2020_2021 %>%
    filter(start <= as.Date("2020-01-01") & vacc > 152) %>%
    mutate(vacc = vacc - 152) %>%
    as.data.frame(),
  failcodes = c("vaccination", "death")
)
saveRDS(fit_vacc_deathComp, "fit_vacc_deathComp.rds")

# subset analyses
fit_vacc_deathComp_subsets <- Cuminc("vacc", "vacc_s",
  data_2020_2021 %>%
    filter(start <= as.Date("2020-01-01") & vacc > 152) %>%
    mutate(vacc = vacc - 152) %>%
    as.data.frame(),
  group = "sex_age",
  failcodes = c("vaccination", "death")
)
saveRDS(fit_vacc_deathComp_subsets, "fit_vacc_deathComp_subsets.rds")

#### use method of RIVM ####
# proportion of all alive on 1 March 2021 with at least 1 dose during 2021
df_vacc_RIVM <- data_2020_2021 %>%
  filter(entry == 0 & srv > as.numeric(as.Date("2021-03-01") - as.Date("2019-12-31"))) %>%
  mutate(
    birthyear = as.numeric(substr(BIRTHDAY, 1, 4)),
    year_group = case_when(
      birthyear < 1931 ~ "<1931",
      birthyear < 1936 ~ "1931-1935",
      birthyear < 1941 ~ "1936-1940",
      birthyear < 1946 ~ "1941-1945",
      birthyear < 1951 ~ "1946-1950",
      birthyear < 1956 ~ "1951-1955",
      T ~ "1956"
    )
  ) %>%
  group_by(year_group) %>%
  summarise(
    max_birthyear = max(birthyear),
    total = n(),
    vacc = length(which(vacc_s == "vaccination")),
    prop = round(vacc / total, 2)
  ) %>%
  mutate(
    prop_RIVM = case_when( 
      grepl("<1931", year_group) ~ 0.89, # from RIVM report
      grepl("31|51", year_group) ~ 0.93, # from RIVM report
      grepl("36|41|46", year_group) ~ 0.94, # from RIVM report
      T ~ 0.91 # from RIVM report
    ))
saveRDS(df_vacc_RIVM, "df_vacc_RIVM.rds")

df_vacc_RIVM %>% mutate(
    correctVacc_perc = prop / prop_RIVM,
    incorrectvacc_perc = (prop_RIVM - prop) / prop_RIVM,
    no_vacc = 1 - prop,
    no_vacc_RIVM = 1 - prop_RIVM,
    incorrectNovacc_perc = (no_vacc - no_vacc_RIVM) / no_vacc
  ) %>%
  print(width = Inf)

df_vacc_RIVM %>% mutate(
  vacc_RIVM = round(total * prop_RIVM),
    vacc_diff = (vacc_RIVM - vacc) / vacc_RIVM,
    no_vacc = total - vacc,
    no_vacc_RIVM = total - vacc_RIVM,
    no_vacc_diff = (no_vacc - no_vacc_RIVM) / no_vacc
  ) %>%
  print(width = Inf) %>%
  select(-max_birthyear)
```

# save objects
```{r save all objects, error=T}
obj <- ls()
for (i in 1:length(obj)) {
  saveRDS(get(obj[i]), paste0("analyses_clean/", obj[i], ".rds"))
}
```
